A Toolbox for Manipulating and Assessing Color Palettes for Statistical Graphics

### 2019-03-11 pkg/colorspace/paper/Toolbox/article.Rmd
feedback from KH
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/colorspace@481 edb9625f-4e0d-4859-8d74-9fd3b1da38cb

  Namely, general strategies for three types of palettes are provided:
  (1)\ Qualitative for coding categorical information, i.e., where no particular
  ordering of categories is available. (2)\ Sequential for coding ordered/numeric information, i.e.,
  going from high to low (or vice versa). (3)\ Diverging for coding ordered/numeric
  information around a central neutral value, i.e., where colors diverge from
  neutral to two extremes.

@@

At the core of the package there are various utilities for computing with
color spaces (as the name conveys). Thus, the package helps to map various three-dimensional
representations of color to each other. A particularly important
mapping is the one from the perceptually-based and device-independent color model
HCL (Hue-Chroma-Luminance) to standard Red-Green-Blue (sRGB) which is the basis for color
specifications in many systems based on the corresponding hex codes (e.g., in HTML but also
in \proglang{R}). For completeness further standard color models are included as well in the package:
`polarLUV()` (= HCL), `LUV()`, `polarLAB()`, `LAB()`, `XYZ()`, `RGB()`, `sRGB()`, `HLS()`,
`HSV()`.

@@

@@

At the core of the \pkg{colorspace} package are various utilities for computing with
color spaces (as the name conveys). Thus, the package helps to map various three-dimensional
representations of color to each other [@color:Ihaka:2003]. A particularly important
mapping is the one from the perceptually-based and device-independent color model
HCL (Hue-Chroma-Luminance) to standard Red-Green-Blue (sRGB) which is the basis for color
specifications in many systems based on the corresponding hex codes (e.g., in HTML but also
in \proglang{R}). For completeness further standard color models are included as well in the package.
The connections are illustrated in Figure\ \ref{fig:color-spaces}.
Color models that are (or try to be) perceptually-based are displayed with circles
and models that are not are displayed with rectangles.

@@

All the \proglang{S}4 classes listed above inherit from a virtual class `color` which are
internally always represented by matrices with three columns (corresponding to
the different three dimensions).

Note that since the inception of the color space conversion tools within \pkg{colorspace}'s
\proglang{C} code by @color:Ihaka:2003 other \proglang{R} tools for this purpose became available.
Notably, base \proglang{R} meanwhile provides `grDevices::convertColor()`
[computed in high-level \proglang{R}, @color:R] and
`farver::convert_colour()` [@color:farver] is based on a \proglang{C++} library. For many basic color
conversion purposes the \pkg{colorspace} package and these alternatives are essentially equally suitable,
though [see the discussion in @color:convertColor]. Moreover, base \proglang{R} also provides
`grDevices::hcl()` for mapping HCL representations to hex codes.

@@

The trajectories for the hue are also rather intuitive and straightforward for the three
different types of palettes (constant vs.\ linear). However, the chroma trajectories are probably most complicated
and least obvious from the examples above. Hence, the exact mathematical equations underlying
the chroma trajectories are given in the following (i.e., using the parameters `c1`, `c2`, `cmax`,
and `p1`, respectively) and are depicted in Figure\ \ref{fig:hcl-trajectories}.
Analogous equations apply for the other two coordinates, though.

@@

The GUIs interface the `qualitative_hcl()`, `sequential_hcl()`, and `diverging_hcl()`
palettes from Section\ \ref{sec:hcl_palettes}. They
allow for interactive modification of the arguments of the respective
palette-generating functions, i.e., starting/ending hue, minimal/maximal chroma,
minimal maximal luminance, and a power transformations that control how
quickly/slowly chroma and/or luminance are changed through the palette.
Subsets of the parameters may not be applicable depending on the type of
palette chosen.

@@

* \pkg{Polychrome} [@color:Polychrome]: qualitative palettes with more "categories". This can generate 
  palettes with the (a large number of) 
  colours spaced far apart from each other in Luv space.
  It also has functions for visualising colour palettes in different ways 
  (https://www.biorxiv.org/content/biorxiv/early/2018/04/18/303883.full.pdf)
* \pkg{roloc} [@color:roloc]: colour conversions, but not between
  numeric colour spaces;  instead, from numeric colour spaces to
  colour names.  
  (https://stattech.wordpress.fos.auckland.ac.nz/2018/01/25/2018-01-generating-colour-names-the-roloc-package-for-r/)
* \pkg{colorscience} [@color:colorscience]: This package is for Colour
Science.  It has data sets, does more complex conversions (including
different chromatic adaptation algorithms).  
So more sophisticated than \pkg{colorspace}
in terms of flexibility and low-level detail, BUT it does not do
palette-generation.
* \pkg{paletteer} [@color:paletteer]: Aggregation of palettes.

>>>>

  General strategies for three types of palettes are provided:
  (1)\ Qualitative for coding categorical information, i.e., where no particular
  ordering of categories is available. (2)\ Sequential for coding ordered/numeric information, i.e.,
  going from high to low (or vice versa). (3)\ Diverging for coding ordered/numeric
  information around a central neutral value, i.e., where colors diverge from
  neutral to two extremes.

@@

At the core of the package there are various utilities for computing with
color spaces [@color:Wiki+Colorspace], as the name conveys. Thus, the package helps to map various three-dimensional
representations of color to each other. A particularly important
mapping is the one from the perceptually-based and device-independent color model
HCL (Hue-Chroma-Luminance) to standard Red-Green-Blue (sRGB) which is the basis for color
specifications in many systems based on the corresponding hex codes (e.g., in HTML but also
in \proglang{R}). For completeness further standard color models are included as well in the package:
`polarLUV()` (= HCL), `LUV()`, `polarLAB()`, `LAB()`, `XYZ()`, `RGB()`, `sRGB()`, `HLS()`,
`HSV()`.

@@

This paper is organized as follows...

@@

At the core of the \pkg{colorspace} package are various utilities for computing with
color spaces (as the name conveys). Thus, the package helps to map various three-dimensional
representations of color to each other [@color:Ihaka:2003]. A particularly important
mapping is the one from the perceptually-based and device-independent color model
HCL (Hue-Chroma-Luminance) to standard Red-Green-Blue (sRGB) which is the basis for color
specifications in many systems based on the corresponding hex codes
[@color:Wiki+Webcolors], e.g., in HTML but also in \proglang{R}.
For completeness further standard color models are included as well in the package.
The connections are illustrated in Figure\ \ref{fig:color-spaces}.
Color models that are (or try to be) perceptually-based are displayed with circles
and models that are not are displayed with rectangles.

@@

All \proglang{S}4 classes for color spaces inherit from a virtual class `color` which are
internally always represented by matrices with three columns (corresponding to
the different three dimensions).

Note that since the inception of the color space conversion tools within \pkg{colorspace}'s
\proglang{C} code by @color:Ihaka:2003 other \proglang{R} tools for this purpose became available.
Notably, base \proglang{R} meanwhile provides `grDevices::convertColor()`
[computed in high-level \proglang{R}, @color:R] and
`farver::convert_colour()` [@color:farver] is based on a \proglang{C++} library. For many basic color
conversion purposes the \pkg{colorspace} package and these alternatives are essentially equally suitable,
though [see the discussion in @color:convertColor]. For more complex conversions, including
different chromatic adaptation algorithms, a more comprehensive color science approach
is implemented in the \proglang{R} package \pkg{colorscience} [@color:colorscience].
Finally, base \proglang{R} also provides
`grDevices::hcl()` for mapping HCL representations to hex codes.

@@

The trajectories for the hue are also rather intuitive and straightforward for the three
different types of palettes (constant vs.\ linear). However, the chroma trajectories are probably most complicated
and least obvious from the examples above. Hence, the exact mathematical equations underlying
the chroma trajectories are given in the following (i.e., using the parameters `c1`, `c2`, `cmax`,
and `p1`, respectively) and are depicted in Figure\ \ref{fig:hcl-trajectories}.
Analogous equations apply for the other two coordinates.

@@

The GUIs interface the `qualitative_hcl()`, `sequential_hcl()`, and `diverging_hcl()`
palettes from Section\ \ref{sec:hcl_palettes}. They
allow for interactive modification of the arguments of the respective
palette-generating functions, i.e., starting/ending hue, minimal/maximal chroma,
minimal maximal luminance, and power transformations that control how
quickly/slowly chroma and/or luminance are changed through the palette.
Subsets of the parameters may not be applicable depending on the type of
palette chosen.

@@

* \pkg{Polychrome} [@color:Polychrome, @color:Polychrome2]: qualitative palettes with more "categories". This can generate 
  palettes with the (a large number of) 
  colours spaced far apart from each other in Luv space.
  It also has functions for visualising colour palettes in different ways 
  (https://www.biorxiv.org/content/biorxiv/early/2018/04/18/303883.full.pdf)
* \pkg{roloc} [@color:roloc,@color:roloc2]: colour conversions, but not between
  numeric colour spaces;  instead, from numeric colour spaces to
  colour names.  
  (https://stattech.wordpress.fos.auckland.ac.nz/2018/01/25/2018-01-generating-colour-names-the-roloc-package-for-r/)
* \pkg{colorscience} [@color:colorscience]: This package is for Colour
Science.  It has data sets, does more complex conversions (including
different chromatic adaptation algorithms).  
So more sophisticated than \pkg{colorspace}
in terms of flexibility and low-level detail, BUT it does not do
palette-generation.
* \pkg{paletteer} [@color:paletteer]: Aggregation of palettes.
